---
title: "Direction check report"
author: "Ramón Gallego"
date: "1/9/2018"
output: html_document
params:
  output_folder:
    value: z


---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=params$output_folder)
```

# Direction report

You have successfully split your libraries and run dada2, and merge the forward and reverse
sequences. But you maybe want to double check that you have all sequences in the right direction. This is a report of that feature.

This runs in bash, mostly. We will source the parameters file, copy the path to the original .1.fastq and .2.fastq, and search if the sequences we obtained in our pipeline (from seqnames.txt) are in the direction they promised.


```{r loading packages, echo=FALSE ,message=FALSE}
library (devtools)
library (tidyverse)
library (Biostrings)
library (digest)



results_search=read_delim("matches.txt", delim = "\t", col_names = c("Primer","File", "nMatches"))
results_search
results_search %>%
 group_by(Primer) %>%
  summarise(Total=sum(nMatches)) %>%
  pull (Total) -> results_spread
names(results_spread) <- results_search %>%
  group_by(Primer) %>%
   summarise(Total=sum(nMatches)) %>% pull (1)

results_spread
direction=NULL

# So here is my solution: If all or 99% of the reads are Fwd - then we assume they are all Fwd and no RC is needed
# If all or 99% of the reads are Reverse Complemented - then I assumy they are all RC and we'll proceed with revcom
# IF there is more than 1% difference then we'll call it mixed and throw a warning message and don't do anything

ifelse(results_spread["Rev_Primer"]== 0 | (results_spread["Fwd_Primer"] / results_spread["Rev_Primer"])>100, yes = direction <- "Fwd",
     ifelse(results_spread["Fwd_Primer"]==0 | (results_spread["Rev_Primer"] / results_spread["Fwd_Primer"])>100, yes = direction <- "Rev", no = direction <- "MIXED"))
direction

if (direction == "MIXED") {

  print(" Seems like some of your reads are in the Fwd direction but others are Reverse Complemented - You should look into that")
  
}

if (direction == "Rev") {

  print(" Seems like your reads are in the Reverse Complemented direction - I'm going to fix that for you ")
# Check if there is a Hash key
if (file.exists("hash_key.csv")){
  print("Found a hash key file")
  Hash_key <- read_csv("hash_key.csv")
  ASV_table <- read_csv("ASV_table.csv")

  Hash_key %>%
    mutate (RRC = as.character(reverseComplement(DNAStringSet(.$Sequence)))) -> Hash_key

  Nhash=NULL

  for (i in 1:nrow(Hash_key)){

    Nhash[i] = digest(Hash_key$RRC[i],algo = "sha1",serialize = F,skip = "auto")

  }  
  Hash_key$NHash <- Nhash
  # Now substitute the Hash in the ASV table
  ASV_table %>%
    left_join(Hash_key %>% select(Hash, NHash)) %>%
    select(sample, Hash = NHash, nReads) -> ASV_table
  ASV_table %>%
    write_csv("New_ASV_table.csv")
  print("Saving a new ASV table")
  # And write the new Hash Key
Hash_key %>%
  select(Hash = NHash, Sequence = RRC) %>%
  write_csv("New_hash_key.csv")
print("Saving a new hash key")
}else {    # In case there is no Hash key file, we assume the user did not want hashes and wanted to keep
          # Raw sequences in the ASV.table

ASV_table <- read_csv("ASV_table.csv")

# Check the Hashes are real sequences

if(sum(str_detect(colnames(ASV_table), "Hash")) == 1){
  print ("Not sure what is going on here: Can't find a hash key but you have hashes")
  knitr::knit_exit()
  }else{
  if(sum(str_detect(ASV_table$Sequence, "[^ATGC]")) > 0){
    print ("You have not-IUPAC characters or ambiguities in your sequences")
    knitr::knit_exit()
  }else{
    ASV_table %>%
      mutate (Sequence = as.character(reverseComplement(DNAStringSet(.$Sequence)))) %>%
      write_csv("New_ASV_table.csv")
    print("Saving a new ASV table")
  }  
}  
}
}

if (direction == "Fwd") {

  print ("All your reads are in the Fwd direction - Congratulations")
}


```
