---
title: "Community analysis on Dada2 data"
author: "Ramón Gallego"
date: "1/12/2018"
output: html_document
params:
  folder: 
    value: x
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=params$folder)
```

## This is a quick report on the output of a MiSeq run

To get a report similar to this, with a newly analised dataset, just check yes on the biodiversity report field of the parameters file.


```{r load packages and files}
fastq.folder
library(tidyverse)
library(vegan)
sample.map<-read_delim(paste0(fastq.folder,"/sample_trans.tmp"),col_names = c("Full_Id", "fastq_header","Sample"),delim = "\t")
metadata.file<-read.csv(paste0(fastq.folder,"/EJP_Halo_1Rep_metadata_sub_fix.csv"),colClasses = "character") %>% 
  as.tibble() %>%
  select(sample_id,Site,Month,Position,Distance)
metadata.file.ostrich <- metadata.file %>% filter (str_detect(sample_id,"Ostrich"))
metadata.file <-   metadata.file %>% filter (!str_detect(sample_id,"Ostrich"))

metadata.file$Distance <-metadata.file$Distance %>% fct_relevel("0","1", "3", "6","10", "15") 

ASV_table<-read_csv((paste0(fastq.folder,"/ASV_table.csv")))

```

## Starting basic stats

Let's analyze the basics of the run you just made.

First we'll look at the number of ASVs, number of reads, and other univariate statistics per sample, and how they average across biological replicates

```{r univariate stats}

ASV_table <- ASV_table %>%
  inner_join(metadata.file, by=c("sample"="sample_id")) %>%
  group_by(sample) %>% mutate("Number_of_ASVs"  = n_distinct(Sequence),
                              "Number_of_Reads" = sum(nReads),
                              "Margalef_index" = round((Number_of_ASVs-1)/log(Number_of_Reads),2)) %>%                            
  group_by(sample,Sequence) %>% mutate ("pi" = nReads/Number_of_Reads) %>%
  
  group_by(sample) %>% mutate("Shannon_diversity" = -sum(pi * log(pi)),
                              "Pielou_equity" = Shannon_diversity/log(Number_of_ASVs)
                               ) 

#EJPs visualization calls for two datasets, along and across, both with bare and eelgrass data for reference
Data.to.plot <- ASV_table %>% filter(Position !="Across") %>% mutate(ds="Along")
Data.to.plot <- ASV_table %>% filter(Position !="Along") %>% mutate(ds="Across") %>% rbind (Data.to.plot)  
Data.to.plot$Distance <- fct_recode(Data.to.plot$Distance, "Eelgrass" = "0", "Bare" = "50")
ggplot(data = Data.to.plot,aes(x = Distance)) + 
  geom_boxplot(aes(y=Pielou_equity))+
  facet_grid(Site~ds)

ggplot(data = Data.to.plot,aes(x = Distance)) + 
  geom_boxplot(aes(y=Shannon_diversity))+
  facet_grid(Site~ds)

ggplot(data = filter(ASV_table,Position !="Across"),aes(x = Distance)) + 
geom_boxplot(aes(y=Shannon_diversity), color= "red")+
  ggtitle("Across coast")+
  facet_grid(Site~.)
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
