---
title: "Dada2_report"
author: "Ramón Gallego"
date: "1/9/2018"
output: html_document
params:
  folder:
    value: x
  hash:
    value: y
  cont:
    value: z
  fastqs:
    value: a
  original:
    value: b

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=params$folder)
```

## Dada2 report

You have successfully split your libraries into a pair (or two) of fastq files per sample. Now let's import the output of demultiplex_both_fastq.sh

First load the packages. And let's update the parameters we need
```{r loading packages, echo=FALSE ,message=FALSE}
library (devtools)
library (tidyverse)
library (stringr)
library (dada2)
library (Biostrings)
library (digest)

#fastq.folder="/Users/rgallego/fastqs_demultiplexed_for_DADA2/demultiplexed_20180108_1539"
sample.map <- read_delim(paste0(params$folder,"/sample_trans.tmp"),col_names = c("Full_Id", "fastq_header","Sample"),delim = "\t")
head (sample.map)

path1 <- params$fastqs
params$original[1]
Sys.setenv(READ1=params$original[1])
Sys.setenv(READ2=params$original[2])
```
Firstly, we'll find the patterns that separate our Fwd, Rev, .1 and .2 files. look at the quality of the .1 and .2 reads
```{r listing files}

F1s <- sort(list.files(path1, pattern="_Fwd.1.fastq", full.names = TRUE))
F2s <- sort(list.files(path1, pattern="_Fwd.2.fastq", full.names = TRUE))
R1s <- sort(list.files(path1, pattern="_Rev.1.fastq", full.names = TRUE))
R2s <- sort(list.files(path1, pattern="_Rev.2.fastq", full.names = TRUE))

# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- str_replace(basename(F1s), "_Fwd.1.fastq","")

#Now use only those that reflect a real sample

good.sample.names<-sample.names[sample.names %in% sample.map$fastq_header]
# Introduce here the biological counterpart of the fastq file

real.sample.name <- sample.map[,3][match(good.sample.names,sample.map$fastq_header),1]

# Now subset the number of files to be looked at

F1sgood<-F1s[sample.names %in% sample.map$fastq_header]
F2sgood<-F2s[sample.names %in% sample.map$fastq_header]
R1sgood<-R1s[sample.names %in% sample.map$fastq_header]
R2sgood<-R2s[sample.names %in% sample.map$fastq_header]

```

A nice idea is to plot the Quality of the F1 reads

```{r qplot1, echo=FALSE}
plotQualityProfile(F1sgood[1:4])
```

The quality should be similar to that of the Reverse .1 reads

```{r plot2, echo=FALSE}
plotQualityProfile(R1sgood[1:4])
```

On the other hand, the quality of the .2 reads should decrease earlier
```{r qplot3, echo=FALSE}
plotQualityProfile(F2sgood[1:4])
```

## Now start with the trimming of each read based on the quality we just plotted
```{r filter and trim}
filt_path <- file.path(params$folder, "/filtered") # Place filtered files in filtered/ subdirectory
filtF1s <- file.path(filt_path, paste0(good.sample.names, "_F1_filt.fastq.gz"))
filtF2s <- file.path(filt_path, paste0(good.sample.names, "_F2_filt.fastq.gz"))
out_Fs <- filterAndTrim(F1sgood, filtF1s, F2sgood, filtF2s, truncLen=c(200,200),
                      maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
                      compress=TRUE, multithread=TRUE) # On Windows set multithread=FALSE

filtR1s <- file.path(filt_path, paste0(good.sample.names, "_R1_filt.fastq.gz"))
filtR2s <- file.path(filt_path, paste0(good.sample.names, "_R2_filt.fastq.gz"))
out_Rs <- filterAndTrim(R1sgood, filtR1s, R2sgood, filtR2s, truncLen=c(200,200),
                      maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
                      compress=TRUE, multithread=TRUE)
```


```{r as tibble}
out_Fs_tibble<-tibble(file=dimnames(out_Fs)[[1]], reads.out=out_Fs[,2])
out_Rs_tibble<-tibble(file=dimnames(out_Rs)[[1]], reads.out=out_Rs[,2])
out_Fs_tibble
```

Now the first crucial step: learning the error rates that would be different for fwd and rev reads

```{r learning errors, echo=T}
errF1 <- learnErrors(filtF1s, multithread=TRUE,verbose = 0)
errF2 <- learnErrors(filtF2s, multithread=TRUE,verbose = 0)
errR1 <- learnErrors(filtR1s, multithread=TRUE,verbose = 0)
errR2 <- learnErrors(filtR2s, multithread=TRUE,verbose = 0)

# Write errors to csv to see if they matter at all
tosave <- list(errF1, errF2, errR1, errR2)

saveRDS(tosave, file = "all.errors.rds")

```

Which we can plot now to see the error rates between transitions of each pair of nt
```{r plotErrors}

plotErrors(errF1, nominalQ = T)
plotErrors(errF2, nominalQ = T)
plotErrors(errR1, nominalQ = T)
plotErrors(errR2, nominalQ = T)


```

## Now go to the dereplication step

```{r dereplication, echo=F,message=FALSE}
derepF1s <- derepFastq(filtF1s, verbose=TRUE)
derepF2s <- derepFastq(filtF2s, verbose=TRUE)
derepR1s <- derepFastq(filtR1s, verbose=TRUE)
derepR2s <- derepFastq(filtR2s, verbose=TRUE)

# Name the derep-class objects by the sample names
rownames(out_Fs) <- names(derepF1s) <- names(derepF2s) <- real.sample.name$Sample

rownames(out_Rs) <- names(derepR1s) <- names(derepR2s) <- real.sample.name$Sample

```

## And finally an inference of the sample composition

```{r dadaing, message=FALSE}
dadaF1s <- dada(derepF1s, err = errF1, multithread = TRUE)
dadaF2s <- dada(derepF2s, err = errF2, multithread = TRUE)
dadaR1s <- dada(derepR1s, err = errR1, multithread = TRUE)
dadaR2s <- dada(derepR2s, err = errR2, multithread = TRUE)

```

## We are ready now to merge reads - using the denoised reads and the derep files.
We will start by saving all files to disk.
```{r saving to disk}
tosave <- list(derepF1s, derepF2s, derepR1s, derepR2s, dadaF1s, dadaF2s, dadaR1s,dadaR2s)

   saveRDS(tosave, file = "tosave.rds")


```
The pipeline breaks if for some reason there are samples that don't have any reads passing filters
(or at least that is the current hypothesis)

Let's try to subset only those samples with more than 1000 reads on each direction

```{r merging pairs}

to.keep.F <- map_lgl(dadaF1s, ~(sum(.x$denoised) > 150)) # Which samples have more than 150 reads passing filters in the Fwd direction
to.keep.R <- map_lgl(dadaR1s, ~(sum(.x$denoised) > 150)) #  Which samples have more than 150 reads passing filters in the Rev direction

to.keep <- (to.keep.F + to.keep.R) == 2 # Keep only those that pass boths filters

mergersF <- mergePairs(dadaF1s[to.keep],
   derepF1s[to.keep],
   dadaF2s[to.keep],
   derepF2s[to.keep],
   verbose = 0)

#Run a for loop that adds the number of unique reads that went into each ASV

for (j in 1:length(mergersF)){

  dadaF1s[[j]]@.Data[[2]] %>% rownames_to_column(var="forward") %>% select("forward", "nunq") ->Fwd
  Fwd$forward<-as.integer(Fwd$forward)
  dadaF2s[[j]]@.Data[[2]] %>% rownames_to_column(var="reverse") %>% select("reverse", "nunq") ->Rev
  Rev$reverse<-as.integer(Rev$reverse)

  mergersF[[j]] <- left_join(mergersF[[j]],Fwd, by="forward") %>% left_join(Rev, by="reverse") %>% mutate(nunq=pmin(nunq.x,nunq.y)) %>% select(-nunq.x,-nunq.y)


}

mergersR <- mergePairs(dadaR1s[to.keep],
   derepR1s[to.keep],
   dadaR2s[to.keep],
   derepR2s[to.keep],
   verbose = 0)

for (j in 1:length(mergersR)){

  dadaR1s[[j]]@.Data[[2]] %>% rownames_to_column(var="forward") %>% select("forward", "nunq") ->Fwd
  Fwd$forward<-as.integer(Fwd$forward)
  dadaR2s[[j]]@.Data[[2]] %>% rownames_to_column(var="reverse") %>% select("reverse", "nunq") ->Rev
  Rev$reverse<-as.integer(Rev$reverse)

  mergersR[[j]] <- left_join(mergersR[[j]],Fwd, by="forward") %>% left_join(Rev, by="reverse") %>% mutate(nunq=pmin(nunq.x,nunq.y)) %>% select(-nunq.x,-nunq.y)

}

```

## Now we have to merge the Forward and Reverse Reads to make a unique object

Step 1 is to create sequence tables for each direction, seqtabF and seqtabR
```{r merging F and R (1)}

seqtabF <- makeSequenceTable(mergersF)

dim(seqtabF)

table(nchar(getSequences(seqtabF)))

seqtabR <- makeSequenceTable(mergersR)

dim(seqtabR)

table(nchar(getSequences(seqtabR)))
```


Step  2 is to reverse complement the reverse reads, and checking how many of the reads in F were present in R (if you get a 0 means we have done something wrong). Lastly, we change the reverse sequences and add their Reverse complement instead
```{r merging F and R (2)}
reversed_sequences<-as.character(reverseComplement(DNAStringSet(colnames(seqtabR))))

summary (colnames(seqtabF) %in% reversed_sequences)

summary (reversed_sequences %in% colnames(seqtabF))

colnames(seqtabR)<-reversed_sequences

```

Step 3 does the actual merging and returns another sequence Table

```{r merging F and R (3)}

seqtab.R.df=as.data.frame(seqtabR)

final.seqtab<-data.frame(row.names = rownames(seqtabF))


for (i in 1:ncol(seqtabF)) {   #for each column of the first dataframe

  current_seq<-colnames(seqtabF)[i]

  if (current_seq %in% colnames(seqtab.R.df)) { # is that column present on the second df?

    final.seqtab[,current_seq]<-seqtabF[,i] + seqtab.R.df[,current_seq] #if yes, the new df has the sum of reads

    seqtab.R.df[,current_seq]<-NULL
    #seqtab.R2.df<- seqtab.R2.df[,-current_seq] #we delete the column from the second df to speed up next search


  } else {        # if the column is not present, then the new df is the value of the first df

    final.seqtab[,current_seq] <- seqtabF[,i]

  }



}
# Now cbind both dataset
final.seqtab <- as.matrix (cbind(final.seqtab,seqtab.R.df))

```

## Now get rid of the chimeras

```{r RemovingChimeras, message=F}

seqtab.nochim <- removeBimeraDenovo(final.seqtab, method="consensus", multithread=TRUE)

dim(seqtab.nochim)


```
## Now it is a good time to check whether the sequences are in the Fwd or Reverse-ReverseComplement direction, we'll use 500 sequences to make up our test


```{r, write colnames}
ifelse(length(colnames(seqtab.nochim))>500, to_write<- sample(colnames(seqtab.nochim),size = 500, replace = F), to_write<-colnames(seqtab.nochim))
#to_write<- sample(colnames(seqtab.nochim),size = 500, replace = F)
write_lines(to_write, path = "seqnames.txt")
getwd()
```



## IF selected, proceed with Hashing: create a hash conversion table and saving files in tidyr format

We are going to keep the info in a tidyr format and save it into a csv file
```{r tidying and writing}

seqtab.nochim.df=as.data.frame(seqtab.nochim)

ASV.file <- file.path(params$folder, "ASV_table.csv")

# Now decide if you want hashing or not

if (grepl ("yes", params$hash, ignore.case = TRUE)) {

  conv_file <-  file.path(params$folder,"hash_key.csv")

  conv_table <- tibble( Hash = "", Sequence ="")

  #hashes <- list(NULL)
  
  map_chr (colnames(seqtab.nochim.df), ~ digest(.x, algo = "sha1", serialize = F, skip = "auto")) -> Hashes
  conv_table <- tibble (Hash = Hashes,
                        Sequence = colnames(seqtab.nochim.df))
  
  colnames(seqtab.nochim.df) <- Hashes

  # for (i in 1:ncol(seqtab.nochim.df)) {   #for each column of the dataframe
  # 
  #   current_seq <-colnames(seqtab.nochim.df)[i]
  # 
  #   current_hash <- digest(current_seq,algo = "sha1",serialize = F,skip = "auto")
  # 
  #   hashes[[i]] <-  current_hash
  # 
  #   conv_table [i,]<-c(current_hash, current_seq)
  # 
  #   colnames(seqtab.nochim.df)[i] <- current_hash
  # 
  # }

  write_csv(conv_table, conv_file)


seqtab.nochim.df %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to="Hash", values_to = "nReads") %>%
  filter (nReads > 0) -> current_asv

write_csv(current_asv, ASV.file)

} else { #What do we do if you don't want hashes: two things - Change the header of the ASV table, write only one file

  seqtab.nochim.df %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to="Sequence", values_to = "nReads") %>%
  filter (nReads > 0) ->  current_asv
  write_csv(current_asv, ASV.file)
}


```

##Track the fate of all reads
```{r output_summary}

getN <- function(x) sum(getUniques(x))
track <- as.data.frame(cbind(out_Fs, out_Rs,
                             sapply(dadaF1s, getN), sapply(dadaR1s, getN),
                             sapply(mergersF, getN),sapply(mergersR, getN),
                             rowSums(seqtabF),rowSums(seqtabR),
                             rowSums(final.seqtab), rowSums(seqtab.nochim)))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input_F", "input_R", "filtered_F","filtered_R",
                     "denoised_F", "denoised_R",
                     "merged_F", "merged_R",
                     "tabled_F", "tabled_R",
                     "tabled_together", "nonchim")

write_csv(track, paste0(params$folder,"/dada2_summary.csv"))

```

## Merge with a previous analysis if selected

If we have chosen to merge our dataset with a previous run, we need to gather that info, and merge properly both datasets.
```{r merging with previous datasets}

if (grepl ("yes", params$cont[1], ignore.case = TRUE)) {

old_hash = read_csv(params$cont[2])

old_asv = read_csv(params$cont[3])

all_hash = union_all(old_hash, conv_table)

new.hashes = setdiff(conv_table,old_hash)

reads.new.hashes <- left_join(new.hashes, current_asv, by = c("Hash" = "Sequence"))

all_asv = bind_rows (old_asv, current_asv)

merging.output = tibble(Date = Sys.Date(), Old_db = params$cont[2], Old_ASV = params$cont[3],
                        Input_folder= params$folder, Nseq_old_db = nrow(old_hash), Nseq_new_input = nrow(conv_table),
                        Nseq_new_db = nrow(all_hash), N_newseq = nrow(new.hashes), Nreads_newseq = sum(reads.new.hashes$nReads) )
ifelse(file.exists(params$cont[4]),
       write_csv(merging.output, params$cont[4], append = T),
       write_csv(merging.output, params$cont[4]) )

write_csv(all_hash, paste0(params$folder, "/Hash_db_",Sys.Date(),".csv"))
write_csv(all_asv, paste0(params$folder, "/ASV_all_",Sys.Date(),".csv"))

}

```
