---
title: "Dada2_report"
author: "Ramón Gallego"
date: "1/9/2018"
output: html_document
params:
  folder:
    value: x
  hash:
    value: y
  cont:
    value: z
  fastqs:
    value: a
  original:
    value: b

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=params$folder)
```

## Dada2 report

You have successfully split your libraries into a fastq file per sample and primer. Now let's import the output of the pipeline and run dada2

First load the packages. And let's update the parameters we need
```{r loading packages, echo=FALSE ,message=FALSE}
library (devtools)
library (tidyverse)
library (stringr)
library (dada2)
library (Biostrings)
library (digest)

#fastq.folder="/Users/rgallego/fastqs_demultiplexed_for_DADA2/demultiplexed_20180108_1539"
# sample.map <- read_delim(paste0(params$folder,"/sample_trans.tmp"),col_names = c("Full_Id", "fastq_header","Sample"),delim = "\t")
# head (sample.map)

path1 <- params$fastqs
#params$original[1]
#Sys.setenv(READ1=params$original[1])

```
Firstly, we'll find the patterns that separate our Fwd, Rev, .1 and .2 files. look at the quality of the .1 and .2 reads
```{r listing files}
# list all original fastq files

original.file <- list.dirs(path = file.path(path1, "noprimers"), recursive = F, full.names = T)



F1s <- sort(list.files(path= original.file, pattern= "*.fastq" , full.names = T, recursive = F))

F1s




# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- str_replace(basename(F1s), ".fastq","")


# Introduce here the biological counterpart of the fastq file

real.sample.name <- sample.map[,3][match(good.sample.names,sample.map$fastq_header),1]



```

A nice idea is to plot the Quality of the F1 reads

```{r qplot1, echo=FALSE}
plotQualityProfile(F1s[2])
```


## Now start with the trimming of each read based on the quality we just plotted
```{r filter and trim}
filt_path <- file.path(path1, "/filtered") # Place filtered files in filtered/ subdirectory
filtF1s <- file.path(filt_path, paste0(sample.names, "_F1_filt.fastq.gz"))

out_Fs <- filterAndTrim(F1s, filtF1s, rm.phix=F,
                      compress=TRUE, multithread=TRUE) # On Windows set multithread=FALSE


```


```{r as tibble}
out_Fs_tibble<-tibble(file=dimnames(out_Fs)[[1]], reads.in = out_Fs[,1], reads.out=out_Fs[,2])

out_Fs_tibble

## Keep only those passing filters

out_Fs_tibble %>% 
  filter (reads.out>10) %>% 
  pull(file) -> passing.filter


filtF1s.passing <- file.path(filt_path, str_replace(passing.filter, ".fastq", "_F1_filt.fastq.gz"))

```

Now the first crucial step: learning the error rates 

```{r learning errors, echo=T}
errF1 <- learnErrors(filtF1s.passing, multithread=TRUE,verbose = 0)


# Write errors to csv to see if they matter at all


saveRDS(errF1, file = "all.errors.rds")

```

Which we can plot now to see the error rates between transitions of each pair of nt
```{r plotErrors}

plotErrors(errF1, nominalQ = T)


```

## Now go to the dereplication step

```{r dereplication, echo=F,message=FALSE}
derepF1s <- derepFastq(filtF1s.passing, verbose=TRUE)

names(derepF1s)


```

## And finally an inference of the sample composition

```{r dadaing, message=FALSE}
dadaF1s <- dada(derepF1s, err = errF1, multithread = TRUE)


```

## We are ready now to merge reads - using the denoised reads and the derep files.
We will start by saving all files to disk.
```{r saving to disk}
tosave <- list(derepF1s,  dadaF1s)

   saveRDS(tosave, file = "tosave.rds")


```
The pipeline breaks if for some reason there are samples that don't have any reads passing filters
(or at least that is the current hypothesis)



Step 1 is to create sequence tables for each direction, seqtabF and seqtabR
```{r Table the results}

seqtabF <- makeSequenceTable(dadaF1s)

dim(seqtabF)

table(nchar(getSequences(seqtabF)))


```



## Now get rid of the chimeras - although it would be good to try vsearch chimera detecting tool

```{r RemovingChimeras, message=F}

seqtab.nochim <- removeBimeraDenovo(seqtabF, method="consensus", multithread=TRUE)

dim(seqtab.nochim)


```



## IF selected, proceed with Hashing: create a hash conversion table and saving files in tidyr format

We are going to keep the info in a tidyr format and save it into a csv file
```{r tidying and writing}

seqtab.nochim.df=as.data.frame(seqtab.nochim)

ASV.file <- file.path(path1, "ASV_table_dada.csv")

# Now decide if you want hashing or not

if (grepl ("yes", params$hash, ignore.case = TRUE)) {

  conv_file <-  file.path(path1,"hash_key_dada.csv")

  conv_table <- tibble( Hash = "", Sequence ="")

  #hashes <- list(NULL)
  
  map_chr (colnames(seqtab.nochim.df), ~ digest(.x, algo = "sha1", serialize = F, skip = "auto")) -> Hashes
  conv_table <- tibble (Hash = Hashes,
                        Sequence = colnames(seqtab.nochim.df))
  
  colnames(seqtab.nochim.df) <- Hashes

  # for (i in 1:ncol(seqtab.nochim.df)) {   #for each column of the dataframe
  # 
  #   current_seq <-colnames(seqtab.nochim.df)[i]
  # 
  #   current_hash <- digest(current_seq,algo = "sha1",serialize = F,skip = "auto")
  # 
  #   hashes[[i]] <-  current_hash
  # 
  #   conv_table [i,]<-c(current_hash, current_seq)
  # 
  #   colnames(seqtab.nochim.df)[i] <- current_hash
  # 
  # }

  write_csv(conv_table, conv_file)


seqtab.nochim.df %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to="Hash", values_to = "nReads") %>%
  filter (nReads > 0) -> current_asv

write_csv(current_asv, ASV.file)

} else { #What do we do if you don't want hashes: two things - Change the header of the ASV table, write only one file

  seqtab.nochim.df %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to="Sequence", values_to = "nReads") %>%
  filter (nReads > 0) ->  current_asv
  write_csv(current_asv, ASV.file)
}


```

##Track the fate of all reads
```{r output_summary}

getN <- function(x) sum(getUniques(x))


track <- as.data.frame(cbind(out_Fs, 
                             sapply(dadaF1s, getN), 
                             rowSums(seqtabF),
                             rowSums(seqtab.nochim)))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input_F", "filtered_F",
                     "denoised_F", 
                     "tabled_F", 
                     "nonchim")

write_csv(track, paste0(params$folder,"/dada2_summary.csv"))

```

## Merge with a previous analysis if selected

If we have chosen to merge our dataset with a previous run, we need to gather that info, and merge properly both datasets.
```{r merging with previous datasets}

if (grepl ("yes", params$cont[1], ignore.case = TRUE)) {

old_hash = read_csv(params$cont[2])

old_asv = read_csv(params$cont[3])

all_hash = union_all(old_hash, conv_table)

new.hashes = setdiff(conv_table,old_hash)

reads.new.hashes <- left_join(new.hashes, current_asv, by = c("Hash" = "Sequence"))

all_asv = bind_rows (old_asv, current_asv)

merging.output = tibble(Date = Sys.Date(), Old_db = params$cont[2], Old_ASV = params$cont[3],
                        Input_folder= params$folder, Nseq_old_db = nrow(old_hash), Nseq_new_input = nrow(conv_table),
                        Nseq_new_db = nrow(all_hash), N_newseq = nrow(new.hashes), Nreads_newseq = sum(reads.new.hashes$nReads) )
ifelse(file.exists(params$cont[4]),
       write_csv(merging.output, params$cont[4], append = T),
       write_csv(merging.output, params$cont[4]) )

write_csv(all_hash, paste0(params$folder, "/Hash_db_",Sys.Date(),".csv"))
write_csv(all_asv, paste0(params$folder, "/ASV_all_",Sys.Date(),".csv"))

}

```
