---
title: "Dada2_report"
author: "Ramón Gallego"
date: "1/9/2018"
output: html_document
params:
  folder: 
    value: x
    
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir=params$folder)
```

## Dada2 report

You have successfully split your libraries into a pair (or two) of fastq files per sample. Now let's import the output of demultiplex_both_fastq.sh  

First load the packages. And let's update the parameters we need
```{r, echo=FALSE ,message=FALSE}
library(devtools)
library(tidyverse)
library(stringr)
library (dada2)
#fastq.folder="/Users/rgallego/fastqs_demultiplexed_for_DADA2/demultiplexed_20180108_1539"
sample.map<-read_delim(paste0(fastq.folder,"/sample_trans.tmp"),col_names = c("Full_Id", "fastq_header","Sample"),delim = "\t")

list.files(path = params$folder)
path1= params$folder
#path1= "/Users/rgallego/fastqs_demultiplexed_for_DADA2/demultiplexed_20180108_1539"
```

Firstly, we'll find the patterns that separate our Fwd, Rev, .1 and .2 files. look at the quality of the .1 and .2 reads
```{r}

F1s <- sort(list.files(path1, pattern="_Fwd.1.fastq", full.names = TRUE))
F2s <- sort(list.files(path1, pattern="_Fwd.2.fastq", full.names = TRUE))
R1s <- sort(list.files(path1, pattern="_Rev.1.fastq", full.names = TRUE))
R2s <- sort(list.files(path1, pattern="_Rev.2.fastq", full.names = TRUE))

# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- str_replace(basename(F1s), "_Fwd.1.fastq","")

#Now use only those that reflect a real sample

good.sample.names<-sample.names[sample.names %in% sample.map$fastq_header]
F1sgood<-F1s[sample.names %in% sample.map$fastq_header]
F2sgood<-F2s[sample.names %in% sample.map$fastq_header]
R1sgood<-R1s[sample.names %in% sample.map$fastq_header]
R2sgood<-R2s[sample.names %in% sample.map$fastq_header]

```

A nice idea is to plot the Quality of the F1 reads

```{r , echo=FALSE}
plotQualityProfile(F1sgood[1:4])
```

The quality should be similar to that of the Reverse .1 reads

```{r , echo=FALSE}
plotQualityProfile(R1sgood[1:4])
```

On the other hand, the quality of the .2 reads should decrease earlier 
```{r , echo=FALSE}
plotQualityProfile(F2sgood[1:4])
```

## Now start with the trimming of each read based on the quality we just plotted
```{r}
filt_path <- file.path(path1, "filtered") # Place filtered files in filtered/ subdirectory
filtF1s <- file.path(filt_path, paste0(good.sample.names, "_F1_filt.fastq.gz"))
filtF2s <- file.path(filt_path, paste0(good.sample.names, "_F2_filt.fastq.gz"))
out <- filterAndTrim(F1sgood, filtF1s, F2sgood, filtF2s, truncLen=c(240,160),
                      maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
                      compress=TRUE, multithread=TRUE) # On Windows set multithread=FALSE
```


```{r}
out_tibble<-tibble(file=dimnames(out)[[1]], reads.out=out[,2])


```

Now the first crucial step: learning the error rates that would be different for fwd and rev reads

```{r, echo=F}
errF1 <- learnErrors(filtF1s, multithread=TRUE)
errF2 <- learnErrors(filtF2s, multithread=TRUE)
```

Which we can plot now to see the error rates between transitions of each pair of nt
```{r}

plotErrors(errF1, nominalQ = T)
plotErrors(errF2, nominalQ = T)

```

## Now go to the dereplication step

```{r, echo=F,message=FALSE}
derepFs <- derepFastq(filtF1s, verbose=TRUE)
derepRs <- derepFastq(filtF2s, verbose=TRUE)
# Name the derep-class objects by the sample names
names(derepFs) <- good.sample.names
names(derepRs) <- good.sample.names
```

## And finally an inference of the sample composition

```{r,message=FALSE}
dadaFs <- dada(derepFs, err = errF1, multithread = TRUE)
dadaRs <- dada(derepRs, err = errF2, multithread = TRUE)
```

## We are ready now to merge reads - using the denoised reads and the derep files

```{r}

mergersF<-mergePairs(dadaFs, derepFs, dadaRs, derepRs, verbose=T)
head(mergersF[[1]])

```

## And we can now construct the sequence abundance table

```{r}

seqtab <- makeSequenceTable(mergersF)

dim(seqtab)

table(nchar(getSequences(seqtab)))

```

## Now get rid of the chimeras

```{r, message=F}

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)

library(vegan)
dis_bc<-vegdist(seqtab.nochim, method = "bray",binary = F, diag=T,upper=T)
rename_samples <- function(map_otu, map_sample){
  map_otu[,1] <- map_sample[,3][match(map_otu[,1], map_sample[,2])]
  return(map_otu)
}


```

We are going to keep the info in a tidyr format
```{r}

seqtab.nochim.df=as.data.frame(seqtab.nochim)
seqtab.nochim.df$sample=rownames(seqtab.nochim.df)

seqs_tidy<-gather(seqtab.nochim.df, key=Sequence, value = nReads, -sample)
inner_join(seqs_tidy,sample.map[,2:3],by = c("sample" = "fastq_header"))
rename_samples(seqtab.nochim.df,sample.map)
```
TODO: I don't like the format of the output: we should 

##Track the fate of all reads
```{r}

getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(mergersF, getN), rowSums(seqtab), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoised", "merged", "tabled", "nonchim")
rownames(track) <- good.sample.names
head(track)

```